A weakly structured stem for human origins in Africa

### 2023-03-06 paper.tex
clarification of stem period

\usepackage[margin=1in]{geometry}
\usepackage{url}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}

\usepackage{authblk}

@@

Third, multiple studies have shown a correspondence between phenotypic differentiation, 
usually assessed with measurements of the cranium, and genetic differentiation among 
human populations and between humans and Neanderthals \citep{Relethford1994-mh,
Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI Section~\ref{sec:morpho_diff}). 
This correspondence potentially allows predictions of our model to be related to the 
fossil record. Some \emph{Homo sapiens} fossils, from Iho Eleru, Nigeria (13ka) 
\citep{Bergstrom2021-iw,Harvati2011-xs}, Ishango, 
Democratic Republic of Congo (20-25ka) \citep{Crevecoeur2016-cl}, and Nazlet Khater, 
Egypt (35-40ka) \citep{Crevecoeur2012-we}, 
have morphological features that may reflect gene flow from archaic hominins. If 
this interpretation is correct, these fossils are unlikely to represent populations 
ancestral to contemporary human populations, because we would have expected to detect 
this gene flow, and it is not present in our models. In our multiple-merger model, we do 
infer late gene flow from one of the stems into West African populations at 25ka, but 
this should not be interpreted as admixture with an archaic hominin lineage as
differentiation between stems remains low. The fossil record of Africa is 
sparse during the time period of the stems, but of the fossils that date to this time period, 
some are fairly similar overall in morphology to contemporary humans (e.g., Omo 1 from 
Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological 
features to contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco 
\citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in morphology to 
have been assigned to species other than \emph{Homo sapiens} (e.g., DH1 from Dinaledi, 
South Africa \citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic 
differences between the stems were comparable to those among contemporary human populations, 
the most morphologically divergent fossils are unlikely to represent branches that 
contributed appreciably to contemporary human ancestries.

>>>>

\usepackage[margin=1in]{geometry}
\usepackage{url}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{authblk}

@@

Third, multiple studies have shown a correspondence between phenotypic differentiation, 
usually assessed with measurements of the cranium, and genetic differentiation among 
human populations and between humans and Neanderthals \citep{Relethford1994-mh,
Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI Section~\ref{sec:morpho_diff}). 
This correspondence potentially allows predictions of our model to be related to the 
fossil record. Some \emph{Homo sapiens} fossils, from Iho Eleru, Nigeria (13ka) 
\citep{Harvati2011-xs}, Ishango, 
Democratic Republic of Congo (20-25ka) \citep{Crevecoeur2016-cl}, and Nazlet Khater, 
Egypt (35-40ka) \citep{Crevecoeur2012-we}, 
have morphological features that may reflect gene flow from archaic hominins. If 
this interpretation is correct, these fossils are unlikely to represent populations 
ancestral to contemporary human populations, because we would have expected to detect 
this gene flow, and it is not present in our models. In our multiple-merger model, we do 
infer late gene flow from one of the stems into West African populations at 25ka, but 
this should not be interpreted as admixture with an archaic hominin lineage as
differentiation between stems remains low. The fossil record of Africa is 
sparse during the time period of the stems \aprcomment{($\gtrsim120$ka)},
but of the fossils that date to this time period, 
some are fairly similar overall in morphology to contemporary humans (e.g., Omo 1 from 
Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological 
features to contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco 
\citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in morphology to 
have been assigned to species other than \emph{Homo sapiens} (e.g., DH1 from Dinaledi, 
South Africa \citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic 
differences between the stems were comparable to those among contemporary human populations, 
the most morphologically divergent fossils are unlikely to represent branches that 
contributed appreciably to contemporary human ancestries.
